---
title: "BRFSS v180 - R2026"
author: EB"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# about

## v180 % Disability: Functional Limitations

### Basic information

- Description
Percentage of adults reporting any of six specific functional limitations.

- What is the numerator?
The numerator is the number of respondents who answered ‘yes’ to at least one of the six following questions:

    * “Are you deaf or do you have serious difficulty hearing?”
    * “Are you blind or do you have serious difficulty seeing, even when wearing glasses?”
    * “Because of a physical, mental, or emotional condition, do you have serious difficulty concentrating, remembering, or making decisions?”
    * “Do you have serious difficulty walking or climbing stairs?”
    * “Do you have difficulty dressing or bathing?”
    * “Because of a physical, mental, or emotional condition, do you have difficulty doing errands alone such as visiting a doctor´s office or shopping?”

- What is the denominator?
The denominator is the total number of adult respondents in a county.

### Raw data access

**Directions**

Two datasets are required

(1) PLACES https://data.cdc.gov/500-Cities-Places/PLACES-County-Data-GIS-Friendly-Format-2025-releas/i46a-9kgh

(2) Disability and Health Data System (DHDS)  https://data.cdc.gov/Disability-Health/Disability-and-Health-Data-System-DHDS-/k62p-6esq/data_preview

PLACES dataset is used to obtain county values: "disability_adjprev", "disability_adj95ci" 

DHDS dataset is used to obtain state and national values: where response = 'Any Disability'

When downloading DHDS dataset, include the following filters:
- YearStart: "2023"
- YearEnd: "2023"
- DataValueTypeID: "AGEADJPREV"     (keep data for age-adjusted prevalence)
- StratificationCategoryID1: "OVERALL"     (keep data for overall population)

Note: for the 2025 release, DHDS state and national data are not available, so the median of counties will be used for state and national values

**Online documentation**

https://www.cdc.gov/places/methodology/index.html

# import packages

```{r}
library(tidyverse)
library(haven)
library(dplyr)

```

# standard county and state fips

- The final dataset will need to be aligned with standard FIPS codes and to include all counties and states.
```{r }
fips <-  bind_rows(
  read_sas("../inputs/county_fips_with_ct_old.sas7bdat") ,
  read_sas( "../inputs/state_fips.sas7bdat"))  %>% 
  arrange(statecode, countycode) %>% 
  glimpse()

df_county_fips <- fips %>% 
  filter(countycode != "000") %>%  
  select(1,2,5) %>% 
  glimpse()

df_state_fips <- fips %>% 
  filter(countycode == "000") %>% 
  select(statecode, countycode, state_abbr = state) %>% 
  glimpse()
```

# raw data

## PLACES 2025 release (2023 data ) - county data

```{r}
raw_dir <- "../raw_data/BRFSS/BRFSS_raw.zip"

places_brfss_county <-  read_csv("../BRFSS/PLACES__County_Data_(GIS_Friendly_Format),_2025_release_20251205.csv")  %>%
  janitor::clean_names() %>% 
  glimpse()
```


# helpers

## get county data from PLACES

```{r}
get_vx_county <- function(df, vnum, var_adj, var_ci, df_fips_cnty = df_county_fips){
  
  vx_county_raw <- df %>% 
    select(StateAbbr, CountyName, CountyFIPS,
           {{ var_adj }}, {{ var_ci }})
  
  vx_county_cleaned <- vx_county_raw %>% 
    mutate({{ var_ci }} := str_remove_all({{ var_ci }}, "[\\(\\)]")) %>% 
    separate({{ var_ci }}, c("vx_cilow", "vx_cihigh"), sep = ",") %>% 
    rename(fipscode = CountyFIPS, vx_rawvalue = {{ var_adj }}) %>% 
    mutate(vx_rawvalue = vx_rawvalue / 100,
           vx_cilow = as.numeric(vx_cilow) / 100,
           vx_cihigh = as.numeric(vx_cihigh) / 100) %>% 
    rename(!!(paste0(vnum, "_rawvalue")) := vx_rawvalue,
           !!(paste0(vnum, "_cilow")) := vx_cilow,
           !!(paste0(vnum, "_cihigh")) := vx_cihigh)
  
  vx_county <- df_fips_cnty %>%
    left_join(vx_county_cleaned,
              by = "fipscode") %>% 
    select(-c(fipscode,StateAbbr,CountyName))
  
  return(vx_county)
  
}
```


## add flag_CT

- This function adds a flag to the data frame for Connecticut counties.
- 'A' = data *available* for a CT county
- 'U' = data *unavailable* for a CT county

```{r}
add_flag_CT <- function(df, vnum, old="U", new = "A"){
  CT_old_lst <- c("001", "003", "005", "007", "009", "011", "013", "015")
  CT_new_lst <- c("110", "120", "130", "140", "150", "160", "170", "180", "190")

  res <- df %>% 
    mutate(flag_CT = case_when(
      statecode == "09" & countycode %in% CT_old_lst ~ old,
      statecode == "09" & countycode %in% CT_new_lst ~ new,
      TRUE ~ NA
    )) %>% 
    rename(!!(paste0(vnum, "_flag_CT")) := flag_CT)
  
  return(res)

}

```


# v180 % Disability: Functional Limitations (new measure in 2025)

## process data

```{r}
v180 <-
    get_vx_county(
      df = places_brfss_county,
      vnum = "v180",
      var_adj = DISABILITY_AdjPrev,
      var_ci = DISABILITY_Adj95CI
    ) %>%
  arrange(statecode, countycode) %>%
  add_flag_CT(vnum = "v180", old = "U", new = "A") %>%
  glimpse()
```

## get state and national data by taking median of counties

```{r}
state_data <- v180 %>%
  group_by(statecode) %>%
  summarise(v180_rawvalue = median(v180_rawvalue, na.rm = TRUE))%>%
  mutate(countycode = "000")

national_data <- v180 %>%
  summarise(v180_rawvalue = median(v180_rawvalue, na.rm = TRUE)) %>%
  mutate(statecode = "00") %>%
  mutate(countycode = "000")

v180_1 <- bind_rows(v180, state_data, national_data)
v180_2 <- v180_1 %>% arrange(statecode, countycode) %>%
  glimpse()
```


##  save data

```{r}
write_csv(v180_2,
          paste0("../measure_datasets/", "v180", "_r2026",
                 ".csv"),
          na = "")
```
